---
title: "FinalProject"
output:
  word_document: default
  html_document: default
---
###########################################################
# This file conducts a time series analysis on vaccinations 
# in the US and forecasts when all adults will receive one.
###########################################################

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###########################################################
# Loading in the data and converting it to a time series.
###########################################################

```{r}
library(tidyverse)
library(dplyr)
library(ggpubr)
library(tseries)
library(astsa)
library(stats)
library(forecast)
library(fpp2)
library(knitr)
library(lubridate)

totaldata<-read.csv("projectdata.csv", skip=3)

data<-subset(totaldata, Date.Type == "Admin", select=c("Date", "Total.Doses.Administered.Daily"))

data$Total.Doses.Administered.Daily <- as.numeric(data$Total.Doses.Administered.Daily)

datafinal <- data %>% group_by(Date=data$Date) %>% summarize(TotalVaccinations = sum(Total.Doses.Administered.Daily))
datafinal<-head(datafinal, -3)
datafinal
vectorvalues<- datafinal$TotalVaccinations

myts <- ts(vectorvalues, start=decimal_date(ymd("2020-12-14")), frequency=7)
```

###########################################################
# First look at the data.
###########################################################

```{r}
# Initially plotting the time series with a best fit line.
plot.ts(myts, ylab="Vaccines Administered", main="Vaccines Administered Per Day")+abline(reg=lm(myts~time(myts)))
```

# Strong trend in the data, with seasonality suspected as well.

```{r}
# Plotting the difference to see if we can make the data stationary.
plot.ts(diff(myts), ylab="Vaccines Administered", main="Vaccines Administered Per Day")

# Since we still have increasing variance in the data, we will take the natural log as well.
plot.ts(diff(log(myts)), main="Difference of the Log of the Time Series")
adf.test(diff(log(myts))) # p-value of <.01 indicated our data is stationary
kpss.test(diff(log(myts)), null="Trend") # p-value of .1 indicates we accept null that is trend stationary


# Since this appears to be trend-stationary, we will now plot the seasonality of the data.
ggseasonplot(diff(log(myts)))
ggsubseriesplot(diff(log(myts)))

```
##########################################################################
# The time series myts has trend, seasonality, and increasing variance.
# The natural log and first difference were taken to remove these aspects.
# The resulting time series still has seasonality.
##########################################################################
# Now we will try forecast methods and evaluate them.
##########################################################################
```{r}
####################################################
# To begin, the seasonal naive model will be used.
####################################################
fitsn <- snaive(myts, h=180) # Residual SD = 0.7978
print(summary(fitsn))
checkresiduals(fitsn)

######################################
# Now we will try the fit ETS method
######################################
fitets <- ets(myts) # Residual SD = 0.3212
print(summary(fitets)) # Model chose multiplicative error, additive trend, multiplicative seasonality
checkresiduals(fitets)

########################################
# Now we will try the fit on ARIMA model
########################################
fitarima <- auto.arima(myts, stepwise=FALSE, approximation=FALSE, trace=TRUE) # 1245
print(summary(fitarima))
checkresiduals(fitarima)

########################################
# Forecast with Seasonal Naive method
########################################
fcstsn <- forecast(fitsn, h=160)
autoplot(fcstsn)

########################################
# Forecast with ETS method
########################################
fcstets <- forecast(fitets, h=160)
autoplot(fcstets)

########################################
# Forecast with ARIMA method
########################################
fcstarima <- forecast(fitarima, h=160)
autoplot(fcstarima)

```
```{r}
# ETS method prediction.
sumtonow<-sum(datafinal$TotalVaccinations)
popvalue = 656400000 - sumtonow
length(which(cumsum(fcstets$mean) <= popvalue))
# The result for ETS is 161, or July 28th
```

```{r}
# Now we will use the ARIMA model to predict.
sumtonow<-sum(datafinal$TotalVaccinations)
popvalue = 656400000 - sumtonow
length(which(cumsum(fcstarima$mean) <= popvalue))
# The result for ARIMA is 147, or August 7th
```
```{r appendix, echo=TRUE, eval=FALSE, ref.label=all_labels()}

```

